---
title: Evaluation of the SV genotypes in trios from the 1000GP
output: github_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, fig.width=10)
```

```{r}
library(dplyr)
library(ggplot2)
library(parallel)
library(gridExtra)
library(knitr)
library(tidyr)
## graph list
ggp = list()
```

## Load genotypes, genotype qualities and trio information

```{r eval=FALSE}
## allele counts
ac.mat = read.table('relkgp.svsite80al.ac.tsv.gz', as.is=TRUE)
ac.mat = as.matrix(ac.mat)

## genotype qualities
gq.mat = read.table('relkgp.svsite80al.gq.tsv.gz', as.is=TRUE)
gq.mat = as.matrix(gq.mat)

## sv types
svs = read.table('svs.relkgp.svsite80al.tsv.gz', as.is=TRUE, header=TRUE) %>%
  select(svsite, type) %>% unique

## pedigree with population/gender information
if(!file.exists('20130606_g1k.ped')){
  download.file('ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/technical/working/20130606_sample_info/20130606_g1k.ped', '20130606_g1k.ped')
}

ped = read.table('20130606_g1k.ped', as.is=TRUE, header=TRUE, sep='\t')

## list mendelian errors
me = c('1_0_0', '2_0_0', ## false "de novo"
       '2_0_1', '2_1_0', '2_2_0', '2_0_2', ## being hom while one parent was hom ref
       '1_2_2', ## being only het when both parents are homs
       '0_2_2', '0_2_0', '0_2_1', '0_0_2', '0_1_2') ## no allele when at least one parent was hom

## minimum GQ thresholds to compute the Mendelian error on
gq.ths = c(0:2, seq(4, 10, 2), seq(15, 30, 5), 40, 50, 75, seq(100, 300, 50))
```

## Number of variants for each GQ threshold

The Mendelian error will be computed for different GQ threshold.
We want to know how many variants each threshold represent.

```{r eval=FALSE}
gq.max.tot = tibble(svsite=rownames(ac.mat), gq.max=apply(gq.mat, 1, max))
gq.max.tot = merge(gq.max.tot, svs)

gq.n = lapply(gq.ths, function(gq.th){
  gq.max.tot %>% group_by(type) %>% summarize(n=sum(gq.max>=gq.th), .groups='drop') %>% mutate(gq.th=gq.th)
}) %>% bind_rows

write.table(gq.n, file='relkgp-gq-nb-variants.tsv', sep='\t', quote=FALSE, row.names=FALSE)
```

## Analyze variants in each trio

```{r eval=FALSE}
## trios that have genotypes/allele counts
trios = ped %>% filter(Paternal.ID %in% colnames(ac.mat),
                       Individual.ID %in% colnames(ac.mat),
                       Maternal.ID %in% colnames(ac.mat))

## loop over each trio
me.df = mclapply(1:nrow(trios), function(ii){
  trio1 = trios[ii,]
  ## allele counts for the offspring_father_mother
  ac = ac.mat[,c(trio1$Individual.ID, trio1$Paternal.ID, trio1$Maternal.ID)]
  ac.v = apply(ac, 1, paste, collapse='_')
  ## maximum genotype quality at each site
  gq = gq.mat[,c(trio1$Individual.ID, trio1$Paternal.ID, trio1$Maternal.ID)]
  gq.max = apply(gq, 1, max)
  ## mendelian error stats for different minimum GQ thresholds
  me.all = tibble(sample=trio1$Individual.ID, svsite=rownames(ac),
                  ac=ac.v, gq.max=gq.max) %>%
    mutate(error=ac%in%me) %>% merge(svs)
  me.s = lapply(gq.ths, function(gq.th){
    me.all %>% filter(gq.max<0 | gq.max>=gq.th) %>%
      group_by(type) %>% mutate(tot=n()) %>% filter(error) %>% 
      group_by(sample, type, ac, tot) %>% summarize(prop=n()/tot[1], .groups='drop') %>%
      arrange(desc(prop)) %>% mutate(gq.th=gq.th)
  }) %>% bind_rows
  ## return results
  return(me.s)
}, mc.cores=12) %>% bind_rows

write.table(me.df, file='relkgp-mendelian-error.tsv', sep='\t', quote=FALSE, row.names=FALSE)
```

## Load pre-computed results

The code above was pre-computed on a larger computing server.
We then load the results:

```{r}
me.df = read.table('relkgp-mendelian-error.tsv.gz', as.is=TRUE, header=TRUE)
gq.n = read.table('relkgp-gq-nb-variants.tsv', as.is=TRUE, header=TRUE)
```

## Mendelian error for each genotype quality threshold

```{r me.gq}
ggp$me = me.df %>% group_by(sample, type, gq.th) %>% summarize(prop=sum(prop)) %>%
  group_by(type, gq.th) %>% summarize(prop.m=median(prop), prop.u=min(prop),
                                      prop.l=max(prop)) %>% 
  ggplot(aes(x=factor(gq.th), y=prop.m, colour=type)) +
  theme_bw() +
  geom_pointrange(aes(ymin=prop.l, ymax=prop.u), position=position_dodge(1)) + 
  geom_point(position=position_dodge(1), alpha=.8) +
  scale_colour_brewer(palette="Set1") + 
  ylab('Mendelian error') + xlab('minimum genotype quality') +
  theme(legend.position=c(.99, .99), legend.justification=c(1,1),
        legend.title=element_blank())
ggp$me

me.df %>% group_by(sample, gq.th, type) %>% summarize(prop=sum(prop)) %>%
  group_by(type, gq.th) %>% summarize(prop=mean(prop)) %>%
  pivot_wider(names_from=type, values_from=prop) %>% 
  kable
```

Range spanned by all samples. Point: median Mendelian error across samples. 



## Most common Mendelian error?

For all the variants (no minimum GQ), what is the configuration that contributes the most to the Mendelian error?

```{r me.common}
me.df %>% filter(gq.th==0) %>% 
  group_by(ac, type) %>% summarize(prop=mean(prop)) %>%
  arrange(desc(prop)) %>% head %>% kable
```

The most common error is `1_0_0` which means heterozygous in the offspring but absent from the parents.
These are likely not real *de novo* variants because we are genotyping known variants not discovering them in those samples.
It's not that surprising that rare false positives would be enriched in this configuration.
Of note, a false negative in the parents could also create this error.

## Total number of variants for each GQ threshold

```{r}
ggp$sv.gq = gq.n %>% group_by(type) %>% mutate(prop=n/n[gq.th==0]) %>%
  ggplot(aes(factor(gq.th), y=prop, colour=type)) +
  geom_line(aes(group=type)) +
  geom_point(size=2, alpha=.8) +
  theme_bw() +
  scale_colour_brewer(palette="Set1") + 
  ylab('proportion of SVs') + xlab('minimum genotype quality') +
  scale_y_continuous(breaks=seq(0,1,.1)) + 
  theme(legend.position=c(.99, .99), legend.justification=c(1,1),
        legend.title=element_blank())
ggp$sv.gq

gq.n %>% group_by(type) %>% mutate(prop=n/n[gq.th==0]) %>%
  arrange(type, gq.th) %>%
  pivot_wider(names_from=type, values_from=c(n,prop)) %>% 
  kable
```

## Figures

```{r fig}
## adds a legend title: a), b), etc
plot_list <- function(ggp.l, gg.names=NULL){
  if(is.null(names(ggp.l))) names(ggp.l) = paste0('g', 1:length(ggp.l))
  if(is.null(gg.names)) gg.names = names(ggp.l)
  lapply(1:length(gg.names), function(ii) ggp.l[[gg.names[ii]]] + ggtitle(paste0('(', LETTERS[ii], ')')))
}

grid.arrange(grobs=plot_list(ggp))

pdf('fig-sv-trio-eval.pdf', 7, 6)
grid.arrange(grobs=plot_list(ggp))
dev.off()
```

